Basics of cubic spline models

### 2021-11-24 text/21/cubic-splines.Rmd
Added natural splines to cubic-splines.Rmd

```{r}
xm <- data.frame(
  c1 =rep(1, length(x)),
  c2 =x,
  c3 =x^2,
  c4 =x^3,
  c5 =(x> 5)*rep(1, length(x)),
  c6 =(x> 5)*(x- 5),
  c7 =(x> 5)*(x- 5)^2,
  c8 =(x> 5)*(x- 5)^3,
  c9 =(x>10)*rep(1, length(x)),
  c10=(x>10)*(x-10),
  c11=(x>10)*(x-10)^2,
  c12=(x>10)*(x-10)^3,
  c13=(x>15)*rep(1, length(x)),
  c14=(x>15)*(x-15),
  c15=(x>15)*(x-15)^2,
  c16=(x>15)*(x-15)^3,
  y=NA
)
xm$y[1:21] <- y
```

```{r}
display_terms <- function(df, x, y, lb, z=y) {
  full_data <- data.frame(x=df[ , x], y=df[ , y], z=df[ , z])
  integer_data <- full_data[1:21, ]
  ggplot(full_data, aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      geom_point(aes(x, z), na.rm=TRUE) +
      xlab(" ") +
      ylab(" ")
}

lb1 <- rep(c("Intercept", "Linear term", "Quadratic term", "Cubic term"), 4)
lb2 <- rep(c("for the full range", "restarted at x=5", "restarted at x=10", "restarted at x=15"), each=4)
lb <- paste(lb1, lb2)
for (j in 1:16) {
  xm %>% display_terms("c2", paste0("c", j), lb[j]) -> cplot
  plot(cplot)
}
```

@@

```{r}
lm(y~c2+c3+c4, data=xm) %>%
  augment(newdata=xm) %>%
  data.frame %>%
  display_terms("c2", ".fitted", "Single cubic fit", "y")
```

@@

```{r}
four_independent_cubics <- 
  lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)
augment(four_independent_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Discontinous cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
four_continuous_cubics <- 
  lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)
augment(four_continuous_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Continous but not smooth cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
four_smooth_cubics <- 
  lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm)
augment(four_smooth_cubics, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y), na.rm=TRUE) +
    ggtitle("Continous and Smooth (1st derivative) splines")

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
spline_model <- 
  lm(y~c2+c3+c4+c8+c12+c16, data=xm)
augment(spline_model, newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Continous and smooth (1st and 2nd derivatives) cubic splines")
```

@@

```{r}
plot(x[1:21], linear_spline[1:21 , 1])
plot(x[1:21], linear_spline[1:21 , 2])
plot(x[1:21], linear_spline[1:21 , 3])
plot(x[1:21], linear_spline[1:21 , 4])
plot(x[1:21], linear_spline[1:21 , 5])

lm(xm$y~linear_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    # geom_point(xm[1:21, ], aes(c2, y)) +
    ggtitle("Linear splines")
```

```{r}
quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
kable(quadratic_spline[1:21, ])
plot(x[1:21], quadratic_spline[1:21 , 1])
plot(x[1:21], quadratic_spline[1:21 , 2])
plot(x[1:21], quadratic_spline[1:21 , 3])
plot(x[1:21], quadratic_spline[1:21 , 4])
plot(x[1:21], quadratic_spline[1:21 , 5])

lm(xm$y~quadratic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y), na.rm=TRUE) + 
    ggtitle("quadratic splines")
```

```{r}
cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
kable(cubic_spline[1:21, ])
plot(x[1:21], cubic_spline[1:21 , 1])
plot(x[1:21], cubic_spline[1:21 , 2])
plot(x[1:21], cubic_spline[1:21 , 3])
plot(x[1:21], cubic_spline[1:21 , 4])
plot(x[1:21], cubic_spline[1:21 , 5])

lm(xm$y~cubic_spline) %>%
  augment(newdata=xm) %>%
  ggplot(aes(c2, .fitted)) +
    geom_line() +
    geom_point(aes(c2, y), na.rm=TRUE) + 
    ggtitle("cubic splines")
```

>>>>

```{r}
xm <- data.frame(
  c1 =rep(1, length(x)),
  c2 =x,
  c3 =x^2,
  c4 =x^3,
  c5 =(x> 5)*rep(1, length(x)),
  c6 =(x> 5)*(x- 5),
  c7 =(x> 5)*(x- 5)^2,
  c8 =(x> 5)*(x- 5)^3,
  c9 =(x>10)*rep(1, length(x)),
  c10=(x>10)*(x-10),
  c11=(x>10)*(x-10)^2,
  c12=(x>10)*(x-10)^3,
  c13=(x>15)*rep(1, length(x)),
  c14=(x>15)*(x-15),
  c15=(x>15)*(x-15)^2,
  c16=(x>15)*(x-15)^3,
  x=x,
  y=NA
)
xm$y[1:21] <- y
```

```{r}
display_terms <- function(df, lb) {
  names(df) <- c("x", "y")
  integer_data <- df[1:21, ]
  ggplot(df, aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      geom_point(data=integer_data, aes(x, y)) +
      xlab(" ") +
      ylab(" ")
}

lb1 <- rep(c("Intercept", "Linear term", "Quadratic term", "Cubic term"), 4)
lb2 <- rep(c("for the full range", "restarted at x=5", "restarted at x=10", "restarted at x=15"), each=4)
lb <- paste(lb1, lb2)

for (j in 1:16) {
  xm %>% 
    select(c(17, j)) %>%
    display_terms(lb[j]) -> cplot
  plot(cplot)
}

@@

```{r}
display_fit <- function(model_fit, lb) {
  model_fit %>%
    augment(newdata=xm) %>%
      ggplot(aes(c2, .fitted)) +
        ggtitle(lb) +
        geom_line() +
        geom_point(aes(c2, y), na.rm=TRUE) +
      xlab(" ") +
      ylab(" ")
}

lm(y~c2+c3+c4, data=xm) %>%
  display_fit("Single cubic fit")
```

@@

```{r}

lm(y~c2+c3+c4+c5+c6+c7+c8+c9+c10+c11+c12+c13+c14+c15+c16, data=xm)  %>%
  display_fit("Discontinous cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 6:8, 10:12, 14:16)])
lm(y~c2+c3+c4+c6+c7+c8+c10+c11+c12+c14+c15+c16, data=xm)  %>%
  display_fit("Continous but not smooth cubic splines")
```

@@

```{r}
kable(xm[1:20, c(1:4, 7:8, 11:12, 15:16)])
lm(y~c2+c3+c4+c7+c8+c11+c12+c15+c16, data=xm) %>%
  display_fit("Continous and Smooth (1st derivative) splines")

@@

```{r}
kable(xm[1:20, c(1:4, 8, 12, 16)])
lm(y~c2+c3+c4+c8+c12+c16, data=xm) %>%
  display_fit("Continous and smooth (1st and 2nd derivatives) cubic splines")
```

@@

```{r}
display_matrix <- function(df, lb) {
  names(df) <- c("x", "y")
  integer_data <- df[1:21, ]
  ggplot(df, aes(x, y)) +
      ggtitle(lb) +
      geom_line() +
      geom_point(data=integer_data, aes(x, y)) +
      xlab(" ") +
      ylab(" ")
}
for (j in 1:dim(linear_spline)[2]) {
  cbind(xm$c2, linear_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Linear spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~linear_spline) %>%
  display_fit("Linear splines")
```

```{r}
quadratic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=2, intercept=TRUE)
kable(quadratic_spline[1:21, ])

for (j in 1:dim(quadratic_spline)[2]) {
  cbind(xm$c2, quadratic_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Quadratic spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~quadratic_spline) %>%
  display_fit("Quadratic splines")
```

```{r}
cubic_spline <- bs(xm$c2, knots=c(5, 10, 15), degree=3, intercept=TRUE)
kable(cubic_spline[1:21, ])
```

```{r}
for (j in 1:dim(cubic_spline)[2]) {
  cbind(xm$c2, cubic_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Cubic spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~cubic_spline) %>%
  display_fit("Cubic splines")
```

```{r}
natural_spline <- ns(xm$c2, knots=c(5, 10, 15), intercept=TRUE)
kable(natural_spline[1:21, ])
```

```{r}
for (j in 1:dim(natural_spline)[2]) {
  cbind(xm$c2, natural_spline[ , j]) %>% 
    data.frame %>%
    display_matrix("Natural spline terms") %>%
    plot
}
```

```{r}
lm(xm$y~natural_spline) %>%
  display_fit("Natural splines")
```
